---
title: "exploratory analysis"
output: html_notebook
---

```{r echo=FALSE}
# set chunk dir to project root dir (and not script location)
knitr::opts_knit$set(root.dir = rstudioapi::getActiveProject())
```

```{r message=FALSE, warning=FALSE}
# libraries
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(glue)
library(scales)
library(ggplot2)
library(tidygraph)
library(ggraph)
```

```{r message=FALSE, warning=FALSE}
# read in transactions dataset
data <- readr::read_csv("data/bs140513_032310.csv" )
```

## data stats
  - **`r scales::comma(nrow(data))`** transactions (records)    
  - **`r ncol(data)`** columns  
  - **(`r c(min(data$step), max(data$step))`**) range of transaction _days_
    (approx 6 months)  
  - **`r scales::comma(length(unique(data$customer)))`** unique customers  
  - **`r scales::comma(length(unique(data$merchant)))`** unique merchants  
  - **`r scales::comma(length(unique(data$category)))`** unique transaction categories  
  - **`r scales::comma(sum(is.na(data)))`** missing values  
  - **`r scales::percent(mean(data$fraud))`** fraud prevalence  
  - **`r scales::dollar(min(data$amount))`** min transaction amount  
  - **`r scales::dollar(mean(data$amount))`** average transaction amount  
  - **`r scales::dollar(max(data$amount))`** max transaction amount  
  
## data-clean up
There are several issues with the data that will require a bit of cleaning.
**Note**: this is not intended to be a data/modeling pre-processing clean up
task, instead, it will make the data appear more _life-like_. That is, `step`
will be turned into a timestamp, and leading ticks will be removed from the
input dataset. Run the `data-updater.R` script to make this happen.  

```{sh eval=FALSE}
# from the project root-directory
Rscript R/data-updater.R
```

# data-exploration
```{r message=FALSE, warning=FALSE}
# read in transactions dataset
transactions <- readr::read_csv("data/transactions.csv")
```

```{r}
# data-summary
trans_summary <- purrr::imap_dfr(transactions, ~ list(
  varname = names(transactions[.y]),
  num_unique = length(unique(.x)),
  pct_unique = round(length(unique(.x))/nrow(transactions), 3), 
  pct_missing = sum(is.na(.x))/nrow(transactions)))

knitr::kable(trans_summary)
```

```{r}
transactions %>% 
  dplyr::group_by(transaction_dt) %>%
  dplyr::summarise_if(is.numeric, mean) %>% 
  qplot(transaction_dt, amount, data=.) +
  geom_line() +
  geom_point() +
  scale_y_continuous(label=scales::dollar) +
  theme_bw() + 
  ggtitle("mean amount spent", "average grouped by transaction_dt")
```

```{r}
transactions %>% 
  dplyr::group_by(transaction_dt, fraud) %>%
  dplyr::summarise_if(is.numeric, mean) %>% 
  dplyr::ungroup() %>%
  dplyr::mutate(fraud = ifelse(fraud == 1, "fraud", "not-fraud")) %>%
  qplot(transaction_dt, amount, data=., color=fraud) +
  geom_line() +
  geom_point() +
  scale_y_continuous(label=scales::dollar) +
  theme_bw() + 
  ggtitle("mean amount spent by fraudulent vs. non-fraudulent charges",
          "average grouped by transaction_dt, fraud")
```

```{r}
transactions %>% 
  dplyr::group_by(transaction_dt, fraud, gender) %>%
  dplyr::summarise_if(is.numeric, mean) %>% 
  dplyr::ungroup() %>%
  dplyr::mutate(fraud = ifelse(fraud == 1, "fraud", "not-fraud")) %>%
  ggplot(aes(transaction_dt, amount, color=gender, alpha=.9), data=.) +
  facet_wrap(~fraud, scales = "free") +
  geom_point() +
  scale_y_continuous(label=scales::dollar) +
  theme_bw() + 
  ggtitle("mean amount spent by gender; fraud vs. non-fraud charges",
          "average grouped by transaction_dt, fraud, gender")
```

```{r}
transactions %>% 
  dplyr::group_by(transaction_dt, fraud, category) %>%
  dplyr::summarise_if(is.numeric, mean) %>% 
  dplyr::ungroup() %>%
  dplyr::mutate(fraud = ifelse(fraud == 1, "fraud", "not-fraud")) %>%
  ggplot(aes(transaction_dt, amount, color=category, alpha=.9), data=.) +
  facet_wrap(~fraud, scales = "free") +
  geom_point() +
  scale_y_continuous(label=scales::dollar) +
  theme_bw() + 
  ggtitle("mean amount spent by category; fraud vs. non-fraud charges",
          "average grouped by transaction_dt, fraud, category")
```
```{r}
transactions %>% 
  dplyr::group_by(transaction_dt, fraud, age) %>%
  dplyr::summarise_if(is.numeric, mean) %>% 
  dplyr::ungroup() %>%
  dplyr::mutate(fraud = ifelse(fraud == 1, "fraud", "not-fraud")) %>%
  ggplot(aes(transaction_dt, amount, color=age, alpha=.9), data=.) +
  facet_wrap(~fraud, scales = "free") +
  geom_point() +
  scale_y_continuous(label=scales::dollar) +
  theme_bw() + 
  ggtitle("mean amount spent by age; fraud vs. non-fraud charges",
          "average grouped by transaction_dt, fraud, age")
```

# network
```{r, eval=FALSE}
gtbl <- transactions %>% 
  dplyr::select(customer, merchant, dplyr::everything())

gtbl <- tidygraph::as_tbl_graph(gtbl)

gtbl
```

